Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-27 Texfiles/5348H.tex
final commit

        Word count: 8336\footnote{This is an estimate that my LaTeX editor gave me.} + 1650 (figure and table word equivalent)
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

The key identifying assumption for DD is the parallel trends assumption. In this
context, it means that consumption differences between the DB, DC and no pension
groups across cohorts would have been the same were it not for the policy change.
Therefore for DD to be invalid, we would need a reason to expect that the
individuals with a DB and no pension who are retiring post 2014 are impacted by
some factor that did not equally impact the DC group. Given that there were no
other pension reforms at this time that differently affected these groups, I
claim that the parallel trends assumption holds.

@@

Theoretically, a null or small change in consumption is evidence for a
bequest motive explaining a lack of annuitisation rather than pessimistic life
expectancies being the cause. In the next, section I solve and simulate a
lifecycle model for individuals in the ELSA dataset. I find that the bequest
motive most closely matches the patterns observed in the empirical models above.

\section{Life cycle theory}

@@

Having found optimal policy functions and value functions, I can now simulate the
change in consumption given different rates of annuitisation versus no
annuitisation. There are multiple benefits to building a model of individual's
decisions. Firstly, we can directly compare what a given individual would have
consumed with what they actually consumed, conditional on our model being
correct. In other words, we have the counterfactual world in which individuals
were not forced to annuitise. Secondly, I can simulate the impact of different
policy changes on the consumption, savings and bequest behaviour of individuals.
For example, I could change the proportion of wealth that will be annuitised and
see the corresponding changes in consumption indicators for all individuals.

@@

Figures \ref{fig:AssetPlot} and \ref{fig:ConsumpPlot} show the retirement path
of consumption and assets for a median retiree. This hypothetical female has
£120,000 of assets at age 65 when she retired in 2012 with a state pension
income of £6000. The dashed lines show consumption and assets with an annuity
that costs half of her wealth. This buys her £3500 of annuity income, which,
given her objective life expectancy is a money's worth of about 90\%.

We can see that the asset path with the bequest motive remains higher for longer
but still decreases, and by her late 90s she has completely run down her
wealth. Interestingly, the asset path with the quickest run down is the standard
lifecycle model using objective life expectancies, which points towards this
individual being optimistic about her life chances. Also rather surprisingly,
the asset paths of all three types are similar without annuities. The paths only
diverge after age 75, at which point the bequest path stays high and the other
paths start to decrease faster. This is contrary to the path with annuities
which show the standard path diverging from the bequest and subjective paths
immediately.

@@

To simulate consumption and asset paths for the individuals in my ELSA dataset.
I round financial wealth and state pension income to the closest point on the
discrete asset and pension grids respectively. I then take the treatment group
and evaluate their consumption in the year of interview with an annuity, given
that they annuitised 50\% of their financial wealth in their retirement year. I
also evaluate consumption without annuitisation. The treatment group did not
need to annuitise their wealth, so I use their predicted consumption from the
lifecycle models with their starting values of assets and pension income taken
from their values in the ELSA dataset. To be clear, the only individuals for
whom I annuitise wealth are those with DC pension pots who retired before 2014.

>>>>

        Word count: 8336\footnote{This is an estimate that my \LaTeX editor gave me.} + 1650 (figure and table word equivalent)
        \vfill
        I confirm that this is entirely my own work and has not previously been submitted for assessment, and I have read and understood the University’s and Faculty’s definition of Plagiarism
    \end{center}
\end{titlepage}
\newpage
\maketitle
\begin{abstract}
    I identify the consumption responses of retirees to the UK's `pensions
    freedom act', a landmark law change that meant defined contribution pension
    holders were no longer forced to annuitise their funds and could instead
    access their pensions in a variety of ways. I use a discontinuity in year of
    retirement and a difference-in-difference approach, harnessing those who
    were unaffected by the policy reform as a natural control group. I then
    solve three sets of lifecycle models, tracking assets from retirement until an
    agent reaches age 110. I simulate their consumption with and without
    annuities, mimicking the policy reform, and match their consumption
    responses to what was seen in the empirical models. I find that a lifecycle
    model with a bequest motive best fits the identified consumption response —
    a result that could inform policymakers in designing optimal pension
    policies for an ageing population.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

The key identifying assumption for DD is the parallel trends assumption. In this
context, it means that consumption differences between the DB, DC and no pension
groups across cohorts would have been the same were it not for the policy change.
Therefore, for DD to be invalid, we would need a reason to expect that the
individuals with a DB and no pension who are retiring post-2014 are impacted by
some factor that did not equally impact the DC group. Given that there were no
other pension reforms at this time that differently affected these groups, I
claim that the parallel trends assumption holds.

@@

Theoretically, a null or small change in consumption is evidence for a
bequest motive explaining a lack of annuitisation rather than pessimistic life
expectancies being the cause. In the next section, I solve and simulate a
lifecycle model for individuals in the ELSA dataset. I find that the bequest
motive most closely matches the patterns observed in the empirical models above.

\section{Lifecycle theory}

@@

Having found optimal policy functions and value functions, I can now simulate the
change in consumption given different rates of annuitisation versus no
annuitisation. There are multiple benefits to building a model of individuals'
decisions. Firstly, we can directly compare what a given individual would have
consumed with what they actually consumed, conditional on our model being
correct. In other words, we have the counterfactual world in which individuals
were not forced to annuitise. Secondly, I can simulate the impact of different
policy changes on the consumption, savings and bequest behaviour of individuals.
For example, I could change the proportion of wealth that will be annuitised and
see the corresponding changes in consumption indicators for all individuals.

@@

Figures \ref{fig:AssetPlot} and \ref{fig:ConsumpPlot} show the retirement path
of consumption and assets for a median retiree. This hypothetical female has
£120,000 of assets at age 65 when she retired in 2012 with a state pension
income of £6,000. The dashed lines show consumption and assets with an annuity
that costs half of her wealth. This buys her £3,500 of annuity income, which,
given her objective life expectancy, is a money's worth of about 90\%.

We can see that the asset path with the bequest motive remains higher for longer
but still decreases, and by the woman's late 90s she has completely run down her
wealth. Interestingly, the asset path with the quickest run down is the standard
lifecycle model using objective life expectancies, which points towards this
individual being optimistic about her life chances. Also rather surprisingly,
the asset paths of all three types are similar without annuities. The paths only
diverge after age 75, at which point the bequest path stays high and the other
paths start to decrease faster. This is contrary to the path with annuities
which show the standard path diverging from the bequest and subjective paths
immediately.

@@

To simulate consumption and asset paths for the individuals in my ELSA dataset,
I round financial wealth and state pension income to the closest point on the
discrete asset and pension grids respectively. I then take the treatment group
and evaluate their consumption in the year of interview with an annuity, given
that they annuitised 50\% of their financial wealth in their retirement year. I
also evaluate consumption without annuitisation. The treatment group did not
need to annuitise their wealth, so I use their predicted consumption from the
lifecycle models with their starting values of assets and pension income taken
from their values in the ELSA dataset. To be clear, the only individuals for
whom I annuitise wealth are those with DC pension pots who retired before 2014.
